---
title: "DAF Input Format for MCView"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DAF Input Format for MCView}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

MCView reads data from DAF (Data Axes Format) objects. This document specifies
**exactly** what your input DAF needs to contain for MCView to work.

A DAF organizes data along named **axes** (e.g., `metacell`, `gene`, `type`).
Each axis has **entries** (names) and can hold **vectors** (one value per entry)
and **matrices** (values at axis intersections).

## Quick Start

The absolute minimum DAF for MCView requires:

1. Three axes: `metacell`, `gene`, `type`
2. One matrix: `UMIs` (metacell × gene)
3. Three vectors: `total_UMIs`, `type` (on metacell), `color` (on type)
4. 2D coordinates: either `x`+`y` or `u`+`v` (on metacell)

Everything else is optional and enables additional tabs/features.

## Required Components

### Axes

| Axis | Description | Example entries |
|------|-------------|-----------------|
| `metacell` | Metacell identifiers | `"M0"`, `"M1"`, ..., `"M2416"` |
| `gene` | Gene identifiers | `"TP53"`, `"GAPDH"`, `"CD4"` |
| `type` | Cell type identifiers | `"T-cell"`, `"B-cell"`, `"Monocyte"` |

### Matrix: UMIs

| Property | Value |
|----------|-------|
| Rows axis | `metacell` |
| Columns axis | `gene` |
| Name | `UMIs` |
| Type | Numeric (non-negative integers) |
| Storage | Sparse or dense |

This is the raw UMI count matrix. MCView computes all derived quantities
(EGC, fold-enrichment, etc.) from this matrix and `total_UMIs`.

### Required Vectors

| Axis | Name | Type | Description |
|------|------|------|-------------|
| `metacell` | `total_UMIs` | numeric | Total UMI count per metacell |
| `metacell` | `type` | character | Cell type assignment (must match `type` axis entries) |
| `type` | `color` | character | Hex color for each cell type (e.g., `"#FF0000"`) |

### 2D Coordinates (at least one pair required)

| Axis | Name | Type | Description |
|------|------|------|-------------|
| `metacell` | `x` | numeric | X coordinate for 2D projection |
| `metacell` | `y` | numeric | Y coordinate for 2D projection |

**Or** alternatively:

| Axis | Name | Type | Description |
|------|------|------|-------------|
| `metacell` | `u` | numeric | U coordinate (alternative to x) |
| `metacell` | `v` | numeric | V coordinate (alternative to y) |

If both pairs exist, MCView uses `x`/`y`.

## Optional Components

### Gene Flags (vectors on `gene` axis)

These boolean vectors enable gene filtering in the UI.

| Name | Type | Description |
|------|------|-------------|
| `is_lateral` | logical | Lateral genes (excluded from certain analyses) |
| `is_noisy` | logical | Noisy genes |
| `is_marker` | logical | Marker genes (enables **Markers** tab auto-detection) |
| `module` | character | Gene module assignment (enables **Gene modules** tab) |

### Metacell Metadata (vectors on `metacell` axis)

| Name | Type | Description |
|------|------|-------------|
| `n_cell` | integer | Number of cells per metacell |
| `mc_col` | character | Per-metacell color override (hex) |
| `top1_gene` | character | Top expressed gene (hover info) |
| `top2_gene` | character | Second top expressed gene |
| `top1_lfp` | numeric | Log fold-enrichment of top gene |
| `top2_lfp` | numeric | Log fold-enrichment of second gene |
| `mc_age` | numeric | Metacell time/age (enables time plots) |

If `top1_gene`/`top2_gene`/`top1_lfp`/`top2_lfp` are missing, MCView computes
them from the UMI matrix on first load. Precomputing saves startup time.

### QC Scalars

| Name | Type | Description |
|------|------|-------------|
| `qc_stats_n_outliers` | integer | Number of outlier cells |
| `qc_stats_n_cells` | integer | Total number of cells |
| `qc_stats_n_umis` | integer | Total UMI count |
| `qc_stats_median_umis_per_metacell` | numeric | Median UMIs per metacell |
| `qc_stats_median_cells_per_metacell` | numeric | Median cells per metacell |

### Gene Correlations

Precomputed gene-gene correlations across metacells. Stored as a custom axis
with associated vectors. If missing, MCView computes correlations on-the-fly
(which is slower for large datasets).

**Axis:** `gg_mc_top_cor` (one entry per correlation pair)

| Name | Type | Description |
|------|------|-------------|
| `gene1` | character | First gene in the correlation pair |
| `gene2` | character | Second gene in the pair |
| `cor` | numeric | Correlation value (-1 to 1) |
| `type` | character | `"pos"` or `"neg"` |

### Metacell Graph (for network visualization)

**Axis:** `metacell_graph` (one entry per edge)

| Name | Type | Description |
|------|------|-------------|
| `from` | character | Source metacell |
| `to` | character | Target metacell |
| `weight` | numeric | Edge weight |
| `graph_name` | character | Graph identifier (default: `"metacell"`) |

### Cell-Level Data (enables **Samples** tab)

**Axis:** `cell` (one entry per cell)

| Name | Type | Required for Samples | Description |
|------|------|---------------------|-------------|
| `metacell` | character | Yes | Metacell assignment per cell |
| `samp_id` | character | Yes | Sample ID per cell |

Additional cell-level vectors are auto-discovered and displayed.

### Atlas Projection (enables **Atlas**, **Query** tabs)

These vectors/matrices on the query DAF enable atlas projection features.
A separate atlas DAF must also be provided.

**Vectors on `metacell`:**

| Name | Type | Description |
|------|------|-------------|
| `projected_type` | character | Projected type from atlas |
| `projected_correlation` | numeric | Correlation with atlas projection |
| `similar` | logical | Whether similar to atlas |
| `atlas_metacell` | character | Most similar atlas metacell |

**Matrices:**

| Rows | Columns | Name | Description |
|------|---------|------|-------------|
| `gene` | `metacell` | `projected_fold` | Projected fold from atlas |
| `metacell` | `gene` | `projected_fraction` | Projected expression fraction |
| `metacell` | `gene` | `corrected_fraction` | Corrected expression fraction |

**Scalars:**

| Name | Type | Description |
|------|------|-------------|
| `projection_weights_json` | character | JSON with query-atlas metacell mapping |
| `query_cell_type_fracs_json` | character | JSON with query cell type fractions |

## Configuration Scalars

You can embed MCView configuration directly in the DAF:

| Name | Type | Description |
|------|------|-------------|
| `mcview_title` | character | Application title |
| `mcview_tabs` | character | Comma-separated list of tabs to show |
| `mcview_excluded_tabs` | character | Comma-separated list of tabs to exclude |
| `mcview_light_version` | logical | Use light version (fewer UI controls) |
| `mcview_about_markdown` | character | About page content (markdown) |
| `mcview_cache_in_daf` | logical | Cache computed data in DAF |
| `metacells_algorithm` | character | Metacells algorithm version |

Configuration priority (highest to lowest):

1. YAML config file passed to `run_app()`
2. DAF scalars (`mcview_*`)
3. Auto-detection based on available data

## Tab Availability

MCView automatically enables tabs based on the data present in the DAF:

| Tab | Requirements |
|-----|-------------|
| **Manifold** | Core requirements only |
| **Genes** | Core requirements only |
| **Diff. Expression** | Core requirements only |
| **Cell types** | Core requirements only |
| **Annotate** | Core requirements only |
| **QC** | Core + QC metadata recommended |
| **Markers** | `gene.is_marker` vector |
| **Gene modules** | `gene.module` vector |
| **Samples** | `cell` axis + `cell.metacell` + `cell.samp_id` |
| **Flow** | Time annotation data in metacell metadata |
| **Atlas** / **Query** | Atlas projection data + separate atlas DAF |

## Validation

Use `validate_daf_for_mcview()` to check your DAF before launching:

```{r, eval = FALSE}
library(MCView)
library(dafr)

daf_obj <- dafr::files_daf("path/to/my_daf")

# Quick check (required components only)
validate_daf_for_mcview(daf_obj)

# Detailed check with verbose output
validate_daf_for_mcview(daf_obj, strict = TRUE, verbose = TRUE)

# Check which tabs will be available
validate_mcview_tabs(daf_obj, verbose = TRUE)
```

## Precomputable Data

For best performance, precompute these fields in your pipeline:

**High priority** (significant startup time savings):

- Gene correlations (`gg_mc_top_cor` axis) -- avoids O(genes²) computation
- Top genes per metacell (`top1_gene`, `top2_gene`, `top1_lfp`, `top2_lfp`)

**Medium priority** (avoids repeated aggregation):

- Gene statistics (`max_umis`, `mean_umis`, `sum_umis` on gene axis)

Use `precompute_mcview_cache()` to add precomputed cache vectors to a writable
DAF:

```{r, eval = FALSE}
daf_obj <- precompute_mcview_cache(daf_obj, verbose = TRUE)
```

## Usage Examples

### Minimal DAF

```{r, eval = FALSE}
library(MCView)
library(dafr)

# Open existing DAF
daf_obj <- dafr::files_daf("path/to/metacells.daf")

# Launch MCView
run_app(daf_obj)
```

### With configuration

```{r, eval = FALSE}
# Launch with custom name and YAML config
run_app(daf_obj,
  dataset_name = "my_experiment",
  config_file = "mcview_config.yaml"
)
```

### Multiple datasets

```{r, eval = FALSE}
daf1 <- dafr::files_daf("path/to/experiment1.daf")
daf2 <- dafr::files_daf("path/to/experiment2.daf")

run_app(list(
  experiment1 = daf1,
  experiment2 = daf2
))
```

### With atlas projection

```{r, eval = FALSE}
query_daf <- dafr::files_daf("path/to/query.daf")
atlas_daf <- dafr::files_daf("path/to/atlas.daf")

run_app(query_daf, atlas = atlas_daf)
```

## Programmatic Contract Access

MCView provides programmatic access to all contract definitions:

```{r, eval = FALSE}
# Get all contracts
contracts <- mcview_all_contracts()

# Get contract for a specific tab
markers_contract <- mcview_tab_contract("Markers")
print_contract(markers_contract)

# Generate full markdown documentation
generate_contract_docs("mcview_contracts.md")
```
