---
title: "Getting started with MCView"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MCView}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Setup

We would start by downloading a PBMC dataset that was pre-processed using the `metacells` python package. 

```{r setup, eval = FALSE}
library(MCView)
```

```{r download, eval=FALSE}
dir.create("raw")
download.file("http://www.wisdom.weizmann.ac.il/~atanay/metac_data/PBMC_processed.tar.gz", "raw/PBMC_processed.tar.gz")
untar("raw/PBMC_processed.tar.gz", exdir = "raw")
```

The above commands end result are two files named "raw/pbmc_metacells.h5ad" and "raw/cluster-colors.csv" which we would import to __MCView__ in the next steps. 

## Create a project

The first step in running __MCView__ is generating a project directory structure: 

```{r, eval=FALSE}
create_project("PBMC")
```

A text editor would be opened with `PBMC/config/config.yaml` file:

```yaml
title: MCView
tabs: ["Manifold", "Genes", "Metacells", "Annotate", "About"] # which tabs to show
help: false # set to true to show introjs help on start
# selected_gene1: Foxc1 # Default selected gene1
# selected_gene2: Twist1 # Default selected gene2
selected_mc1: 1 # Default selected metacell1 
selected_mc2: 2 # Default selected metacell2
```

We can edit some of the parameters if we want, or add parameters per dataset, see the [architecture](https://tanaylab.github.io/MCView/articles/Architecture.html) vignette for a full description of the config parameters. 


## Import 

Next, we would import the PBMC dataset to the project we created. A project can contain multiple dataset, and switching between them can be done from the right sidebar. 

This would pre-process the metacell dataset in order to view it in the shiny app: 

```{r, eval=FALSE}
import_dataset(project = "PBMC", 
               dataset = "PBMC163k", 
               anndata_file = "raw/metacells.h5ad", 
               cell_type_field =  "cluster",
               cell_type_colors_file = "raw/cluster-colors.csv")
```

The most important field is the `anndata` field which points to the h5ad file we downloaded. 

You can see that we also specified a field in the `h5ad` object that has pre-computed cluster assignments per metacell, and the path of a file which contains color assignment per cluster, but these where optional, as __MCView__ can generate initial clustering and coloring automatically. 

Note that the import part might take a few minutes, depending (mostly) on the number of metacells. 

__MCView__ supports also importing datasets from the old R `metacell` package, see `import_dataset_metacell1` for details.

## Run the app

```{r, eval=FALSE}
run_app(project = "PBMC", launch.browser = TRUE)
```

A browser window would be opened with the app. 

You can also specify the port or host, or do not launch the browser window, e.g.: 

```{r, eval=FALSE}
run_app(project = "PBMC", port = 5555, host = "127.0.0.1", launch.browser = FALSE)
```

## Update annotations

After working a bit on the initial metacell model, we would usually want to update the default dataset annotations with the ones we created using __MCView__. This can be done by:

1. Pressing the "export" button in the upper left of "Annotate" screen and saving the file. 
2. Running: 

```{r, eval=FALSE}
update_metacell_types("PBMC", "PBMC163k", "/path/to/metacell_types_file")
```

Where "/path/to/metacell_types_file" is the path of the exported file. 

The same can be done for cell type colors, which has a different export button: 

```{r, eval=FALSE}
update_cell_type_colors("PBMC", "PBMC163k", "/path/to/cell_type_colors_file")
```

You can now rerun the app and the types/colors would be updated. 

Note that if you changed cell type __names__ you should update both `cell_type_colors` and `metacell_types`, as otherwise metacells with the old name would appear with a "(Missing)" cell type tag. 

## Deploy

Create a deployment ready bundle by running: 

```{r, eval=FALSE}
create_bundle(project = "PBMC", path = getwd(), name = "PBMC_bundle")
```

You can then upload the bundle to [shinyapps.io](https://docs.rstudio.com/shinyapps.io/getting-started.html#deploying-applications) by running: 

```{r, eval=FALSE}
rsconnect::deployApp(appDir = file.path(getwd(), "PBMC_bundle"))
```

Or to any shiny-server hosting service by uploading the "PBMC_bundle" directory to the service. 

## Docker 

See the [docker](https://tanaylab.github.io/MCView/articles/Docker.html) vignette for instructions of using the docker image. 