---
title: "Getting started with MCView"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MCView}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Setup

We would start by downloading a PBMC dataset that was pre-processed using the `metacells` python package. 

```{r setup, eval = FALSE}
library(MCView)
```

```{r download, eval=FALSE}
dir.create("raw")
download.file("http://www.wisdom.weizmann.ac.il/~atanay/metac_data/PBMC_processed.tar.gz", "raw/PBMC_processed.tar.gz")
untar("raw/PBMC_processed.tar.gz", exdir = "raw")
```

The above commands end result are two files named "raw/pbmc_metacells.h5ad" and "raw/cluster-colors.csv" which we would import to __MCView__ in the next steps. 

## Create a project

The first step in running __MCView__ is generating a project directory structure: 

```{r, eval=FALSE}
create_project("PBMC", title = "PBMC")
```

A text editor would be opened with `PBMC/config/config.yaml` file:

```yaml
title: PBMC
tabs: ["Manifold", "Genes", "Markers", "Diff. Expression", "Cell types","Annotate", "About"] # which tabs to show
help: false # set to true to show introjs help on start
# selected_gene1: Foxc1 # Default selected gene1
# selected_gene2: Twist1 # Default selected gene2
selected_mc1: 1 # Default selected metacell1 
selected_mc2: 2 # Default selected metacell2
```

The configuration file was generated using `create_project` parameters, but we can edit them if we want, or add parameters per dataset, see the [architecture](https://tanaylab.github.io/MCView/articles/Architecture.html) vignette for a full description of the config parameters. 


## Import 

Next, we would import the PBMC dataset to the project we created. A project can contain multiple dataset, and switching between them can be done from the right sidebar. 

This would pre-process the metacell dataset in order to view it in the shiny app: 

```{r, eval=FALSE}
import_dataset(project = "PBMC", 
               dataset = "PBMC163k", 
               anndata_file = "raw/metacells.h5ad", 
               cell_type_field =  "cluster",
               cell_type_colors_file = "raw/cluster-colors.csv")
```

The most important field is the `anndata` field which points to the h5ad file we downloaded. 

You can see that we also specified a field in the `h5ad` object that has pre-computed cluster assignments per metacell, and the path of a file which contains color assignment per cluster, but these where optional, as __MCView__ can generate initial clustering and coloring automatically. 

Note that the import part might take a few minutes, depending (mostly) on the number of metacells. If you see that it takes too long - set `calc_gg_cor` to `FALSE` in order to skip calculating correlation between all genes. This would save significant amount of import time but would make this feature unavailable in the app. 

__MCView__ supports also importing datasets from the old R `metacell` package, see `import_dataset_metacell1` for details.

## Run the app

```{r, eval=FALSE}
run_app(project = "PBMC", launch.browser = TRUE)
```

A browser window would be opened with the app. 

You can also specify the port or host, or do not launch the browser window, e.g.: 

```{r, eval=FALSE}
run_app(project = "PBMC", port = 5555, host = "127.0.0.1", launch.browser = FALSE)
```

## Update annotations

After working a bit on the initial metacell model, we would usually want to update the default dataset annotations with the ones we created using __MCView__. This can be done by:

1. Pressing the "export" button in the upper left of "Annotate" screen and saving the file. 
2. Running: 

```{r, eval=FALSE}
update_metacell_types("PBMC", "PBMC163k", "/path/to/metacell_types_file")
```

Where "/path/to/metacell_types_file" is the path of the exported file. 

The same can be done for cell type colors, which has a different export button: 

```{r, eval=FALSE}
update_cell_type_colors("PBMC", "PBMC163k", "/path/to/cell_type_colors_file")
```

You can now rerun the app and the types/colors would be updated. 

Note that if you changed cell type __names__ you should update both `cell_type_colors` and `metacell_types`, as otherwise metacells with the old name would appear with a "(Missing)" cell type tag. 

## Metadata

You can load numeric metadata fields for each metacell using the `metadata` parameter in `import_dataset` command:

```{r, eval=FALSE}
import_dataset(project = "PBMC", 
               dataset = "PBMC163k", 
               anndata_file = "raw/metacells.h5ad", 
               cell_type_field =  "cluster",
               cell_type_colors_file = "raw/cluster-colors.csv",
               metadata = "raw/metadata.csv"
               )
```

The format is a data frame (or a delimited filename) with a column named `metacell` and the annotation fields. The metadata fields can be either numeric or categorical. 

You can use the `metadata_colors` parameter to set the breaks and colors for each numerical metadata field, and color for each category for categorical fields. 

If you want to change the metadata fields or colors after the import, you can use the  `update_metadata` and `update_metadata_colors` functions:

```{r, eval=FALSE}
update_metadata(project = "PBMC", 
               dataset = "PBMC163k", 
               metadata = "new_metadata.csv")
update_metadata_colors(project = "PBMC", 
               dataset = "PBMC163k", 
               metadata_colors = new_metadata_colors)               
```

You can generate metadata per metacell from cell metadata using the `cell_metadata_to_metacell` and `cell_metadata_to_metacell_from_h5ad` functions. 

## Deploy

Create a deployment ready bundle by running: 

```{r, eval=FALSE}
create_bundle(project = "PBMC", path = getwd(), name = "PBMC_bundle")
```

You can then upload the bundle to [shinyapps.io](https://docs.rstudio.com/shinyapps.io/getting-started.html#deploying-applications) by running: 

```{r, eval=FALSE}
rsconnect::deployApp(appDir = file.path(getwd(), "PBMC_bundle"))
```

Or to any shiny-server hosting service by uploading the "PBMC_bundle" directory to the service. 

> Note that you might need to set your hosting service to allow higher memory than the default - __MCView__ keeps the metacell matrix in-memory and therefore needs around 1GB of RAM for small datasets such as PBMC, but up to 2-4GB for large datasets such as MOCA. 

## Docker 

See the [docker](https://tanaylab.github.io/MCView/articles/Docker.html) vignette for instructions of using the docker image. 