---
title: "Getting started with MCView"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MCView}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

## Setup

First, let's load the MCView package:

```{r setup, eval = FALSE}
library(MCView)
```

For this tutorial, we'll use a PBMC dataset that was pre-processed using the `metacells` python package.

```{r download, eval=FALSE}
dir.create("raw")
download.file("http://www.wisdom.weizmann.ac.il/~atanay/metac_data/PBMC_processed.tar.gz", "raw/PBMC_processed.tar.gz")
untar("raw/PBMC_processed.tar.gz", exdir = "raw")
```

This gives us two files: "raw/metacells.h5ad" (the metacell data) and "raw/cluster-colors.csv" (color assignments for cell types).

## Import Dataset

The most common way to start with MCView is to directly import your dataset:

```{r, eval=FALSE}
import_dataset(
    project = "PBMC", # This will create the project if it doesn't exist
    dataset = "PBMC",
    anndata_file = "raw/metacells.h5ad",
    cell_type_field = "type"
)
```

This single command:
1. Creates a project directory structure if it doesn't exist
2. Pre-processes the metacell dataset for viewing in the app
3. Adds an `app.R` file, making the project deployment-ready

The `anndata_file` points to the h5ad file we downloaded, and `cell_type_field` specifies the field in the h5ad object that contains pre-computed cluster assignments.

> **Note**: Import may take a few minutes depending on the number of metacells. If it's too slow, set `calc_gg_cor = FALSE` to skip calculating gene correlations. This saves significant import time but makes this feature unavailable in the app.

> **Tip**: Run `compute_for_mcview` in the `metacells` python package before importing to enable all MCView features.

MCView also supports importing datasets from the old R `metacell` package with `import_dataset_metacell1`.

## Run the app

Once import is complete, you can run the app:

```{r, eval=FALSE}
run_app(project = "PBMC", launch.browser = TRUE)
```

This opens a browser window with the app. You can customize the port, host, and browser launch:

```{r, eval=FALSE}
run_app(project = "PBMC", port = 5555, host = "127.0.0.1", launch.browser = FALSE)
```

If you're already in the project directory, simply run:

```{r, eval=FALSE}
run_app(launch.browser = TRUE)
```

## Update annotations

After working with the initial metacell model, you might want to update the annotations:

1. Export your changes by pressing the "export" button in the upper left of the "Annotate" screen 
2. Update the annotations with:

```{r, eval=FALSE}
update_metacell_types("PBMC", "PBMC", "/path/to/metacell_types_file")
```

Where "/path/to/metacell_types_file" is the path to your exported file.

To only update cell type colors:

```{r, eval=FALSE}
update_cell_type_colors("PBMC", "PBMC", "/path/to/cell_type_colors_file")
```

## Working with metadata

You can add metadata for each metacell during import:

```{r, eval=FALSE}
import_dataset(
    project = "PBMC",
    dataset = "PBMC",
    anndata_file = "raw/metacells.h5ad",
    metadata = "raw/metadata.csv"
)
```

The metadata should be a data frame (or file) with a column named `metacell` and additional annotation fields, which can be numeric or categorical.

Use `metadata_colors` to customize colors for each field.

To update metadata after import:

```{r, eval=FALSE}
update_metadata(
    project = "PBMC",
    dataset = "PBMC",
    metadata = "new_metadata.csv"
)

update_metadata_colors(
    project = "PBMC",
    dataset = "PBMC",
    metadata_colors = new_metadata_colors
)
```

You can generate metacell metadata from cell metadata using `cell_metadata_to_metacell` or `cell_metadata_to_metacell_from_h5ad`.

## Deploy your app

MCView projects are deployment-ready by default. The `app.R` file added during import means you can deploy your app by simply copying the entire project directory to a Shiny server.

### Direct Deployment

Deploy to [shinyapps.io](https://docs.rstudio.com/shinyapps.io/getting-started.html#deploying-applications):

```{r, eval=FALSE}
rsconnect::deployApp(appDir = "PBMC")
```

Or deploy to any Shiny server by copying the "PBMC" directory to the server.

### Advanced Deployment (Optional)

For additional customization (such as including MCView code in your bundle):

```{r, eval=FALSE}
create_bundle(
    project = "PBMC",
    path = getwd(),
    name = "PBMC_bundle",
    self_contained = TRUE # Includes MCView code in the bundle
)

# Then deploy
rsconnect::deployApp(appDir = file.path(getwd(), "PBMC_bundle"))
```

> **Note**: MCView keeps the metacell matrix in memory and requires around 1GB of RAM for small datasets like PBMC, and up to 2-4GB for larger datasets. Adjust your hosting service settings accordingly.

## Advanced: Manual project creation

If you want to customize your project configuration before importing data, you can manually create a project first:

```{r, eval=FALSE}
create_project("PBMC", title = "PBMC")
```

This opens a text editor with `PBMC/config/config.yaml`:

```yaml
title: PBMC
tabs: ["QC", "Manifold", "Genes", "Markers", "Gene modules", "Diff. Expression", "Cell types","Annotate", "About"]
help: false
# selected_gene1: Foxc1
# selected_gene2: Twist1
selected_mc1: 1
selected_mc2: 2
```

After editing, proceed with `import_dataset` as shown earlier.

See the [architecture](https://tanaylab.github.io/MCView/articles/Architecture.html) vignette for a complete description of configuration parameters.

## Docker 

See the [docker](https://tanaylab.github.io/MCView/articles/Docker.html) vignette for instructions on using the Docker image.
